---
title: "COVID-19 - Vaccine efficiency/effectiveness - screening method - overall"
author: "olivier.duperrex@unisante.ch"
date: "version 2021-09-08"
output:
  html_document:
  df_print: paged
fig_caption: yes
toc: no
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
bibliography: references.bib
---


# Aim

-   Calculate vaccine effectiveness approximation using the Farrington
    method [@farrington1993]

# What are we talking about?

-   The so-called Screening Method is very simple and cost-effective to
    have a quick assesment of the vaccine effectiveness in a population

-   It compares the proportion of vaccinated in cases and in the
    population

-   The surveillance data on vaccination coverage must be reliable

-   It is better to stratify by age groups when the coverage varies
    between them

-   Further information on limitations of the method can be found in
    [@ecdc2018] and [@who2021]

# Setting up the scene

```{r setup}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = NA, include = TRUE, results='hide')
# avoid scientific notation
options(scipen=999)


## you are here -----------------------------------------------------
# rm(list = ls())
here::here()

```

## Definitions

-   VE = vaccine efficacy (in controlled research studies) or vaccine
    effectiveness (in real public health world)

+-----------------------+-----------------------------+-----------------+
| Abbreviation          | en                          | fr              |
+=======================+=============================+=================+
| VE                    | vaccine efficacy (in        | efficacité      |
|                       | controlled research         | vaccinale       |
|                       | studies) or vaccine         | étudiée ou      |
|                       | effectiveness (in real      | réelle =\> VE   |
|                       | public health world) =\> VE | est 1 - rapport |
|                       | is 1 -- odds of vaccination | des *cotes* des |
|                       | in cases compared to        | vaccinés parmi  |
|                       | population                  | les cas comparé |
|                       |                             | àla population  |
+-----------------------+-----------------------------+-----------------+
| PCV                   | proportion of cases         | proportion des  |
|                       | vaccinated                  | cas vaccinés    |
+-----------------------+-----------------------------+-----------------+
| PPV                   | proportion of population    | proportion de   |
|                       | vaccinated                  | la population   |
|                       |                             | vaccinée        |
+-----------------------+-----------------------------+-----------------+

If you don't know about this method, please look at [01_VE_nomogram](01_VE_nomogram.html).


## Select language for labels in tables and graphs
If you use this script you might have to modify some of the labels below so they correspond to your reality.


```{r language}

language_chosen <- "en"
# language_chosen <- "fr"



if (language_chosen == 'fr') {
  ## labels fr ---------------------------------
  title <- "Canton factice - COVID-19"
  
  label_y <- "PCV = proportion des cas vaccinés"
  label_x <- "PPV = proportion de la population vaccinée"
  
  label_agegr_gen <-  "Groupe d'âge"
  label_total_pop <- "Population pour ce groupe d'âge (N)"
  label_total_vaccinated_2doses_14d_N <-
    "Personnes complètement vaccinées (N)"
  label_total_cases_N <- "Nouveaux cas (N)"
  # label_cases_14d_vaccinated_2doses_14d_N <- "Cas vaccinés (N)"
  
  label_PPV_2doses_14d <-
    "Proportion de la population vaccinée (PPV)"
  label_PCV_2doses_14d  <-  "Proportion des cas vaccinés (PCV)"
  label_VE_2doses_14d <-  "Efficacité vaccinale réelle (VE)"
  
  
  caption_g <-  "Source : Données factices pour illustration"
  
  subtitle_t1 <-
    "Approximation de l'efficacité vaccinale réelle selon les générations, par la méthode de Farrington"
  
  footnote_t1_1 <-
    "Méthode de dépistage = méthode cas-population. Compare la part des vaccinés dans la population à celle des cas, en utilisant les rapports de cotes (Odds ratios)"
  # footnote_t1_2  <-
  #   "Nombre de nouveaux cas depuis le 27.1.2021 (modification formulaire OFSP)"
  footnote_t1_3 <- "Nouveaux cas 14j ou plus après la 2e dose"
  footnote_t1_4 <- "Avec 2 doses depuis 14j ou plus"
  
  period_1 <- "depuis le 27.1.2021"
  footnote_total_cases_t1 <-
    paste0("Nombre de nouveaux cas ",
           period_1,
           " (modification formulaire OFSP)")
  subtitle_part_1 <-
    "Comparaison des nouveaux cas chez les personnes complètement vaccinées <br> et chez les non ou partiellement vaccinées,
selon les générations "
  
  footnote_per10k <- "Pour 10'000 personnes"
  
  
  
  label_cases_14d_vaccinated_2doses_14d_N <-
    "Cas chez personnes vaccinées (N)"
  label_cases_nonvaccinated_14d <-
    "Cas chez personnes non ou partiellement vaccinées (N)"
  label_total_nonvaccinated_14d <-
    "Personnes non ou partiellement vaccinées (N)"
  label_incidence_cases_vaccinated_2doses_14d_per10k  <-
    'Incidence cas chez les vaccinés (pour 10 mille)'
  label_incidence_cases_nonvaccinated_14d_per10k  <-
    'Incidence cas chez les personnes non vaccinées  (pour 10 mille)'
  
  label_RR <-
    "Risque relatif de devenir un cas chez les personnes non ou partiellement vaccinées comparé aux personnes complètement vaccinées - si RR>1, la vaccination protège"
  label_RR_lower <- 'RR IC95% inf'
  label_RR_upper <- 'RR IC95% sup'
  label_RR_inverse <-
    "Risque relatif de devenir un cas des personnes complètement vaccinées comparé aux personnes non ou partiellement vaccinées - si RR<1, la vaccination protège"
  
  label_vacc_2d_14j  <-
    "Personnes complètement vaccinées (2 doses depuis 14 jours ou plus)"
  
  
  label_y_p02 <- 'Incidence de cas (pour 10 mille)'
  subtitle_part_p02 <-
    "Nouveaux cas pour 10'000 personnes - selon les générations "
  
  label_vacc_2d_14j_fully <- 'Personnes vaccinées complètement'
label_vacc_2d_14j_notfully <- 'Personnes non vaccinées ou partiellement'


} else {
  ## labels en ---------------------------------
  title <- "Fictional canton - COVID-19"
  
  label_y <- "PCV = proportion of cases vaccinated"
  label_x <-  "PPV = proportion of population fully vaccinated"
  
  label_agegr_gen <-  "Age group"
  label_total_pop <- "Population by age group (N)"
  label_total_vaccinated_2doses_14d_N <-
    "Persons fully vaccinated (N)"
  label_total_cases_N <- "New cases (N)"
  label_cases_14d_vaccinated_2doses_14d_N <-
    "New cases despite full vaccination (N)"
  
  label_PPV_2doses_14d <-
    "Proportion of population fully vaccinated (PPV)"
  label_PCV_2doses_14d  <-
    "Proportion of cases fully vaccinated (PCV)"
  label_VE_2doses_14d <-  "Vaccine effectiveness (VE)"
  
  caption_g <-  "Source : Fictional data for illustration"
  
  subtitle_t1 <-
    "Approximation de vaccine effectiveness by generations, with the Farrington method"
  footnote_t1_1 <-
    "Screening method =  case-population method. Compare the proportion of vaccinated between population and cases, by using Odds ratios"
  # footnote_t1_2  <-
  # "Number of new cases since le 27.1.2021 (modification formulaire OFSP)"
  footnote_t1_3 <- "New cases 14d or more after the 2nd dose"
  footnote_t1_4 <- "Vaccinated with 2 doses 14d age or more"
  
  period_1 <- "since 27.1.2021"
  footnote_total_cases_t1 <-
    paste0("Number of new cases ",
           period_1,
           " (modification of FOPH declaration)")
  subtitle_part_1 <-
    "Comparison of new cases in fully vaccinated persons (2 doses 14d ago or more) <br> and in the not-fully vaccinated persons,
by generations "
  
  label_RR <-
    "RR : Relative risk to be a case in not-fully vaccinated persons (no vaccine, 1 unique dose or 2 doses less than 14j) compared to fully vaccinated persons (2 doses 14d ago or more) - if RR>1, vaccination protects"
  label_RR_lower <- 'RR CI95% low'
  label_RR_upper <- 'RR CI95% high'
  footnote_per10k <- "For 10'000 persons"
  label_cases_nonvaccinated_14d <-
    "Cases in persons not-fully vaccinated (N)"
  label_total_nonvaccinated_14d <-
    "Persons not fully vaccinated (N)"
  label_incidence_cases_vaccinated_2doses_14d_per10k  <-
    'Incidence of new cases in fully vaccinated persons (pour 10 mille)'
  label_incidence_cases_nonvaccinated_14d_per10k  <-
    'Incidence of new cases in not-fully vaccinated persons (pour 10 mille)'
  label_RR_inverse <-
    "Relative risk to be a case in fully vaccinated persons (2 doses 14d ago or more) compared to not-fully vaccinated persons (no vaccine, 1 unique dose or 2 doses less than 14j) - if RR<1, vaccination protects"
  
  label_vacc_2d_14j  <-
    "Persons fully vaccinated (2 doses depuis 14 jours ou plus)"
  label_y_p02 <- "Incidence of new cases (per 10'000)"
  subtitle_part_p02 <-
    "New cases per 10'000 persons - selon les générations "
  
  label_vacc_2d_14j_fully <- 'Persons fully vaccinated'
  label_vacc_2d_14j_notfully <- 'Persons not-fully vaccinated'

}

## . p02_bis ---------------------------------------------------------
label_x_p02 <- label_vacc_2d_14j

subtitle_p02 <-
  gt::md(
    paste0(
      subtitle_part_p02,
      '**(',
      period_1,
      ')<br>**'
    )
  )

```

## The equations

```{r equations}

VE_equation <- "VE = 1 - (PCV / (1 - PCV)) / (PPV / (1 - PPV))"
PCV_equation <- "PCV = (PPV - (PPV * VE)) / (1 - (PPV * VE))"

# caption_table : will be used as caption
caption_table <- VE_equation
```

## Load libraries

```{r load libraries}
pacman::p_load(
  data.table,   # managing data
  ggplot2,      # create plots
  ggpmisc,      # plot tweaking - i.e. insert a table in the plot
  ggrepel,      # nice labels on plots
  gt,           # nice tables
  huxtable,     # nice tables
  magrittr,     # for the pipe %>% and the assignement pipe %<>%
  scales        # for percent in graph axis
  # sjmisc
)

```

## Local functions

```{r local functions}
## local functions --------------------------------------------------
## all functions might not be used in this repository but are kept for illustration

source(here::here('code', '00_functions.R'), encoding = 'UTF-8')

```

## Load data


The data used in this script is a fictional data set, inspired by the reality of a canton at one point. You can update the file with your own data (or read your own file) and run this script.


```{r load data}
## load data --------------------------------------------------------
## data_raw_overall.xlsx is a fictional data set
file_to_import <- here::here('data', 'data_raw_overall.xlsx')

data_raw_overall <-
  rio::import(file_to_import) %>%       # rio will select the appropriate package to import
  setDT()     %>%                       # create a data.table
  sjlabelled::var_labels(               # add labels using quasiquotation (double exclamation mark)
    agegr_gen = !!label_agegr_gen,
    total_pop = !!label_total_pop,
    total_vaccinated_2doses_14d_N = !!label_total_vaccinated_2doses_14d_N,
    total_cases_N = !!label_total_cases_N,
    cases_14d_vaccinated_2doses_14d_N = !!label_cases_14d_vaccinated_2doses_14d_N
  )

str(data_raw_overall)

## for more info on use of quasiquotation please read https://strengejacke.github.io/sjlabelled/articles/quasiquotation.html 

```

## Setup plots and tables

```{r setup plots and tables}
## theme for huxtables
options('huxtable.knit_print_df_theme' = theme_article)
options('huxtable.autoformat_number_format' = list(numeric = "%5.2f"))



## themes and elements for plots ---------------------------------------
theme_set(sjPlot::theme_sjplot2(base_size = 10))
theme_update(
  legend.title = element_blank(),
  legend.position = "right",
  strip.background = element_rect(fill = "white"),
  plot.subtitle = ggtext::element_markdown()
)


## Prepare some elements for plots ----------------------------------
# out_png4ppt : output folder for png ---
# needs to be modified according to your own project structure
out_png4ppt <- here::here('output', 'png4ppt') 


caption_table <- caption_g

caption_g_bis <- paste0('
', caption_g)

```

# Generations

-   We use generations rather then multiples of 10 or 20 because they are more meaningful

-   This takes into accounts the observation period of 14d
    post-vaccination

-   The variable *day_14_post_second_vac_date* was created in the
    recoding step and is used to filter the dataset
    


```{r t1_vaccine_effectiveness_by_generations}


## . add total by columns ----

t1_vaccine_effectiveness_by_generations <- data_raw_overall %>% janitor::adorn_totals()


## . PPV and PCV ----
## PPV = proportion of population vaccinated
## PCV = proportion of cases vaccinated
t1_vaccine_effectiveness_by_generations[, `:=`(
  PPV_2doses_14d = formattable::percent(total_vaccinated_2doses_14d_N / total_pop, 1),
  PCV_2doses_14d = formattable::percent(cases_14d_vaccinated_2doses_14d_N / total_cases_N, 1)
)]

## . VE = vaccine effectiveness ----

t1_vaccine_effectiveness_by_generations[, VE_2doses_14d := 1 - (PCV_2doses_14d / (1 - PCV_2doses_14d)) / (PPV_2doses_14d / (1 - PPV_2doses_14d))]

## . modify order of columns ----
new <- c("agegr_gen",
         "total_pop" ,
         "total_vaccinated_2doses_14d_N",
         "PPV_2doses_14d",
         "total_cases_N" ,
         "cases_14d_vaccinated_2doses_14d_N",
         "PCV_2doses_14d",
         "VE_2doses_14d")

setcolorder(t1_vaccine_effectiveness_by_generations, new)


## . add variable labels ----
## need to reapply them
t1_vaccine_effectiveness_by_generations  %<>%
  sjlabelled::var_labels(
    agegr_gen = !!label_agegr_gen,
    total_pop = !!label_total_pop,
    total_vaccinated_2doses_14d_N = !!label_total_vaccinated_2doses_14d_N,
    PPV_2doses_14d = !!label_PPV_2doses_14d,
    total_cases_N = !!label_total_cases_N,
    cases_14d_vaccinated_2doses_14d_N = !!label_cases_14d_vaccinated_2doses_14d_N,
    PCV_2doses_14d = !!label_PCV_2doses_14d,
    VE_2doses_14d = !!label_VE_2doses_14d
  )



## save table as xlsx
name_xlsx <-
  here::here(
    'output',
    'xlsx',
    't1_vaccine_effectiveness_by_generations.xlsx'
  )

writexl::write_xlsx(t1_vaccine_effectiveness_by_generations, name_xlsx)

## save table as RData
filename <-   here::here('output', 'RData', 't1_vaccine_effectiveness_by_generations.RData')
save(t1_vaccine_effectiveness_by_generations,  file = filename)

```



```{r t1_vaccine_effectiveness_by_generations_gt}
# . gt ----

# subtitle_t1 <- "Approximation de l'efficacité vaccinale réelle selon les générations, par la méthode de Farrington"

table_name <- 't1_vaccine_effectiveness_by_generations.png'

# names(t1_vaccine_effectiveness_by_generations)

## generate table with labels
t1_vaccine_effectiveness_by_generations_gt <-
  t1_vaccine_effectiveness_by_generations %>%
  gt() %>%
  tab_header(title = title,
             subtitle = subtitle_t1) %>%
  cols_label(
    agegr_gen = label_agegr_gen,
    total_pop = label_total_pop,
    total_vaccinated_2doses_14d_N = label_total_vaccinated_2doses_14d_N,
    PPV_2doses_14d = label_PPV_2doses_14d,
    total_cases_N = label_total_cases_N,
    cases_14d_vaccinated_2doses_14d_N = label_cases_14d_vaccinated_2doses_14d_N,
    PCV_2doses_14d = label_PCV_2doses_14d,
    VE_2doses_14d = label_VE_2doses_14d

  ) %>%

  fmt_number(columns = c(2:3, 5:6),
             decimals = 0,
             sep_mark = "'") %>%

  tab_footnote(footnote = footnote_t1_1,
               locations =  cells_title(groups = "subtitle")) %>%

  tab_footnote(footnote = footnote_total_cases_t1,
               locations = cells_column_labels(columns = total_cases_N)) %>%
  tab_footnote(
    footnote = footnote_t1_3,
    locations = cells_column_labels(columns = cases_14d_vaccinated_2doses_14d_N)
  ) %>%

  tab_footnote(
    footnote = footnote_t1_4,
    locations = cells_column_labels(
      columns = c(
        total_vaccinated_2doses_14d_N,
        PPV_2doses_14d,
        PCV_2doses_14d,
        VE_2doses_14d
      )
    )) %>%

  tab_footnote(footnote = VE_equation,
  locations = cells_column_labels(columns = VE_2doses_14d)) %>%
      tab_source_note(source_note = caption_table) %>%

      tab_options(data_row.padding = px(2)
      )



# t1_vaccine_effectiveness_by_generations_gt


## save gt table as png
t1_vaccine_effectiveness_by_generations_gt %>% gtsave(table_name,
                     path = out_png4ppt)




```


# Incidence rates amongst fully vaccinated and others

```{r t2_RR_nonvac_vacc_generations_global}
## . t2_RR_nonvac_vacc_generations_global  ----

subtitle <-
  md(
    paste0(
      subtitle_part_1,
      '**(',
      period_1,
      ')**'
    )
  )


table_name <- 't2_RR_nonvac_vacc_generations_global.png'

## t2_RR_nonvac_vacc_generations_global ---
names(t1_vaccine_effectiveness_by_generations)

t2_RR_nonvac_vacc_generations_global <-
  t1_vaccine_effectiveness_by_generations[, .(
    agegr_gen,
    total_pop,
    total_vaccinated_2doses_14d_N,
    PPV_2doses_14d,
    total_cases_N ,
    cases_14d_vaccinated_2doses_14d_N
  )]


multiplicator  <-  10 ^ 4

t2_RR_nonvac_vacc_generations_global[, incidence_cases_vaccinated_2doses_14d_per10k := round(
  multiplicator * (
    cases_14d_vaccinated_2doses_14d_N / total_vaccinated_2doses_14d_N
  ),
  1
)]

t2_RR_nonvac_vacc_generations_global[, cases_nonvaccinated_14d := total_cases_N - cases_14d_vaccinated_2doses_14d_N]

t2_RR_nonvac_vacc_generations_global[, total_nonvaccinated_14d := total_pop - total_vaccinated_2doses_14d_N]

t2_RR_nonvac_vacc_generations_global[, incidence_cases_nonvaccinated_14d_per10k :=  round(multiplicator * (cases_nonvaccinated_14d / total_nonvaccinated_14d), 1)][]

## approach :
## # non-exposed = vaccinated
## # exposed = non-vaccinated

t2_RR_nonvac_vacc_generations_global[, c('RR', 'RR_lower', 'RR_upper',
       'AR', 'AR_lower', 'AR_upper') := add_RR_AR(
         x2 = cases_14d_vaccinated_2doses_14d_N, # non-exposed = vaccinated
         n2 = total_vaccinated_2doses_14d_N,       # non-exposed = vaccinated
         x1 = cases_nonvaccinated_14d,       # exposed = non-vaccinated
         n1 = total_nonvaccinated_14d,       # exposed = non-vaccinated
         rounding = 0
       ),
   keyby = agegr_gen]

## . !!! pb with ARR need to be checked !!! ----
t2_RR_nonvac_vacc_generations_global[, c('AR', 'AR_lower', 'AR_upper') := NULL]

t2_RR_nonvac_vacc_generations_global[, RR_inverse := round(1/RR, 3)][]

## . add labels ----
# foo5b <- copy_labels(foo5b, foo5)
sjlabelled::get_label(t2_RR_nonvac_vacc_generations_global)

# names(t2_RR_nonvac_vacc_generations_global)
t2_RR_nonvac_vacc_generations_global <-
  t2_RR_nonvac_vacc_generations_global %>%
  sjlabelled::var_labels(
    agegr_gen = !!label_agegr_gen,
    total_pop = !!label_total_pop,
    total_vaccinated_2doses_14d_N = !!label_total_vaccinated_2doses_14d_N,
    PPV_2doses_14d = !!label_PPV_2doses_14d,
    cases_14d_vaccinated_2doses_14d_N = !!label_cases_14d_vaccinated_2doses_14d_N,
    cases_nonvaccinated_14d = !!label_cases_nonvaccinated_14d,
    total_nonvaccinated_14d = !!label_total_nonvaccinated_14d,
    incidence_cases_vaccinated_2doses_14d_per10k  = !!label_incidence_cases_vaccinated_2doses_14d_per10k,
    incidence_cases_nonvaccinated_14d_per10k  = !!label_incidence_cases_nonvaccinated_14d_per10k,
    RR = !!label_RR,
    RR_lower = !!label_RR_lower,
    RR_upper = !!label_RR_upper,
    # AR = !!'Différence de risque',
    # AR_lower = !!'AR IC95% inf',
    # AR_upper = !!'AR IC95% sup',
    RR_inverse = !!label_RR_inverse
  )

# sjlabelled::get_label(t2_RR_nonvac_vacc_generations_global)

    

## get rid of the inverse RR
t2_RR_nonvac_vacc_generations_global[, RR_inverse:=NULL]

new_col_order <-
  c(
    "agegr_gen",
    "total_pop",
    "total_vaccinated_2doses_14d_N",
    "total_nonvaccinated_14d",
    "PPV_2doses_14d" ,
    "total_cases_N",
    "cases_14d_vaccinated_2doses_14d_N",
    "cases_nonvaccinated_14d" ,
    "incidence_cases_vaccinated_2doses_14d_per10k" ,
    "incidence_cases_nonvaccinated_14d_per10k",
    "RR",
    "RR_lower",
    "RR_upper"
  )

setcolorder(t2_RR_nonvac_vacc_generations_global, new_col_order)

## generate table with labels ---
t2_RR_nonvac_vacc_generations_global_gt <-
  t2_RR_nonvac_vacc_generations_global %>%
  gt() %>%
  tab_header(title = title,
             subtitle = subtitle) %>%
  cols_label(
    agegr_gen = label_agegr_gen,
    total_pop = label_total_pop,
    total_cases_N = label_total_cases_N,
    total_vaccinated_2doses_14d_N = label_total_vaccinated_2doses_14d_N,
    PPV_2doses_14d = label_PPV_2doses_14d,
    cases_14d_vaccinated_2doses_14d_N = "Cas chez personnes vaccinées (N)",
    cases_nonvaccinated_14d = "Cas chez personnes Autres (N)",
    total_nonvaccinated_14d = "Personnes Autres (N)",
    incidence_cases_vaccinated_2doses_14d_per10k  = 'Incidence cas chez les personnes vaccinées  (pour 10 mille)',
    incidence_cases_nonvaccinated_14d_per10k  = 'Incidence cas chez les personnes Autres  (pour 10 mille)',
    RR = "RR",
    RR_lower = 'RR IC95% inf',
    RR_upper = 'RR IC95% sup'
  ) %>%

  fmt_number(
    columns = c(2:4, 6:8),
    decimals = 0,
    sep_mark = "'"
  ) %>%

  tab_footnote(footnote = footnote_total_cases_t1,
               locations = cells_column_labels(columns = total_cases_N)) %>%
  tab_footnote(
    footnote = "Nouveaux cas 14j ou plus après la 2e dose",
    locations = cells_column_labels(columns = cases_14d_vaccinated_2doses_14d_N)
  ) %>%

  tab_footnote(footnote = "Vaccinée = avec 2 doses depuis 14j ou plus",
               locations = cells_column_labels(columns = c(total_vaccinated_2doses_14d_N,
                                                           PPV_2doses_14d,
                                                           incidence_cases_vaccinated_2doses_14d_per10k))) %>%
  tab_footnote(footnote = "Autres = aucun vaccin, 1 seule dose ou 2 doses depuis moins de 14j",
               locations = cells_column_labels(columns = c(cases_nonvaccinated_14d,
                                                           total_nonvaccinated_14d,
                                                           incidence_cases_nonvaccinated_14d_per10k))) %>%
  tab_footnote(
    footnote =
      label_RR,
    locations = cells_column_labels(columns = c(RR, RR_lower, RR_upper))) %>%
  tab_source_note(source_note = caption_table) %>%

  tab_options(data_row.padding = px(2)
  )



t2_RR_nonvac_vacc_generations_global_gt



t2_RR_nonvac_vacc_generations_global_gt %>% gtsave(table_name, path = out_png4ppt,
                 vwidth = 1200, vheight = 600)




name_xlsx <-
  here::here(
    'output',
    'xlsx',
    't2_RR_nonvac_vacc_generations_global.xlsx'
  )

writexl::write_xlsx(t2_RR_nonvac_vacc_generations_global, name_xlsx)


filename <-   here::here('output', 'RData', 't2_RR_nonvac_vacc_generations_global.RData')
save(t2_RR_nonvac_vacc_generations_global,  file = filename)
```


## Table 
```{r t2_bis_RR_nonvac_vacc_generations_global}
## . t2_bis_RR_nonvac_vacc_generations_global - incidence with CI ---------------------------------------
## reshape and add vacc_2d_14j --



## .. t2_vacc_2d_14j : subset of t7 ----
t2_vacc_2d_14j <-
  t2_RR_nonvac_vacc_generations_global[, .(vacc_2d_14j = label_vacc_2d_14j_fully,
                                           agegr_gen,
                                           total = total_vaccinated_2doses_14d_N,
                                           cases = cases_14d_vaccinated_2doses_14d_N)]

## .. t2_nonvacc_2d_14j : subset of t7 ----
t2_nonvacc_2d_14j <-
  t2_RR_nonvac_vacc_generations_global[, .(vacc_2d_14j = label_vacc_2d_14j_notfully,
                                           agegr_gen,
                                           total = total_nonvaccinated_14d,
                                           cases = cases_nonvaccinated_14d)]

t2_bis_RR_nonvac_vacc_generations_global <- rbind(t2_vacc_2d_14j, t2_nonvacc_2d_14j)


t2_bis_RR_nonvac_vacc_generations_global[, c(
  'inc_per10K',
  'conf_low_per10K',
  'conf_high_per10K'
) := add_prop_test(
  cases,
  total,
  multiplicator = 10 ^ 4,
  rounding = 1
), keyby = .(vacc_2d_14j, agegr_gen)]


# t2_bis_RR_nonvac_vacc_generations_global[, vacc_2d_14j := relevel(as.factor(vacc_2d_14j), ref = 'Oui')]

t2_bis_RR_nonvac_vacc_generations_global %<>%
    sjlabelled::var_labels(
      vacc_2d_14j = label_vacc_2d_14j
  )


name_xlsx <-
  here::here(
    'output',
    'xlsx',
    't2_bis_RR_nonvac_vacc_generations_global.xlsx'
  )

writexl::write_xlsx(t2_bis_RR_nonvac_vacc_generations_global, name_xlsx)

```


## Plots
```{r p02_bis_incidence_cas_vacc2doses_14d_autre_generations}

nudge_x <- 0.2


p02_bis <-
  ggplot(
    t2_bis_RR_nonvac_vacc_generations_global[!(agegr_gen == 'Total'),],
    aes(
      x = agegr_gen,
      y = inc_per10K ,
      ymin = conf_low_per10K ,
      ymax = conf_high_per10K,
      label = round(inc_per10K, 0),
      col = agegr_gen
    )
  ) +
  # ylim(0, 100) +
  scale_y_continuous(
    limits = c(0, 1200),
    breaks = seq(0, 1200, 200),
    labels = comma_format(big.mark = "'", accuracy = 1)
  ) +
  geom_pointrange() +
  # coord_flip() +
  labs(
    title = title,
    subtitle = subtitle,
    x = '',
    y = label_y_p02,
    caption = caption_g_bis
  ) +
  geom_hline(yintercept = 0) +
  scale_x_discrete(labels = NULL, position = 'top') +
  # scale_x_discrete(guide = guide_axis(n.dodge = 3)) +
  # scale_x_discrete(guide = guide_axis(angle = 45)) +
  facet_grid(. ~ vacc_2d_14j) +
  ggrepel::geom_text_repel(
    nudge_x = nudge_x,
    min.segment.length = Inf,
    show.legend = FALSE  # Don't display "a" in the legend.
  ) +
  theme(
    panel.spacing = unit(3, "lines"),
    text = element_text(size = 12),
    strip.placement = 'top',
    strip.background = element_blank()
  ) +
  guides(fill = guide_legend(title = "",
                             override.aes = aes(label = "")))

p02_bis

plot_name <- paste0("p02_bis_incidence_cas_vacc2doses_14d_autre_generations.png")

ggsave(plot_name, path = out_png4ppt, width = 10, height = 6, dpi = 600)

```
